---
title: "Proportional hazards model"
format: html
editor: visual
---

## ================================================================

## Survival Analysis

## Proportional hazards model

## ================================================================

```{r}
load("Ahda_RLab3.RData")
library(survival)
```

## A data set on patients with multiple myeloma -------------------------------

## ----------------------------------------------------------------------------

```{r}
str(multmyel)
head(multmyel)
summary(multmyel)
Hmisc::hist.data.frame(multmyel, nclass = 10)
```

## The Surv object

## ---------------

```{r}
myesurv <- with(multmyel, Surv(time, status))
```

## Two survival plots (drawn with function survplot\[rms\])

## ------------------------------------------------------

```{r}
library(rms)
#quartz(width = 16)             # Under Mac: quartz(); under Linux: x11()
par(mfrow = c(1, 2), font = 2, font.lab = 4, font.axis = 2, las = 1,
    oma = c(0, 0, 1, 0), mar = c(5, 5, 4, 2))
survplot(npsurv(myesurv ~ sex, multmyel), xlab = "Months", col = 1:2,
         lwd = 3, ylab = expression(bolditalic(hat(S)(t))), time.inc = 5)
title("Survival functions according to sex")
survplot(npsurv(myesurv ~ BJP, multmyel), xlab = "Months", col = 1:2,
         lwd = 3, ylab = expression(bolditalic(hat(S)(t))),
         label.curves = FALSE, time.inc = 5, n.risk = TRUE)
title("Survival functions according to Bence-Jones protein")
legend("top", c("BJP: Yes", "BJP: No"), col = 1:2, lwd = 3, bty = "n")
```

## Fit of Cox models----- -----------------------------------------------------

additative - needed for linearity in log!

BJP = protein (biomarker) \## ----------------------------------------------------------------------------

```{r}
cox1 <- coxph(myesurv ~ sex + BJP, multmyel)
summary(cox1)
```

HR = 0.9432 = e\^-0.05852 = e\^beta 6% lower instentaneous risk (hazard) for male (same proteins) but CI includes 1 negative beta (coefficient) is to be interpreted differently from parametric survival function (where pos coef lowers risk)

baseline of hazard (intercept) is not shown and not necesaary for HR

## Including the interaction between both variables

## ------------------------------------------------

```{r}
cox1 <- update(cox1, ~ . + sex:BJP)
 summary(cox1)
```

## Exercise:

## Compute the model-based hazard ratios associated with the absence of the

## BJ protein.

female: HR = 1.5980

male: HR = e\^(0.4688+0.1131) !!!!

```{r}
exp(0.4688+0.1131)
```

## ------------------------------------------------------------------------

## The Cox model with BUN, HB, age, sex, and BJP and remove interaction

## ---------------------------------------------

```{r}
cox2 <- update(cox1, ~ . + BUN + HB + age - sex:BJP, multmyel)
print(summary(cox2), signif.stars = FALSE)
```

see the tests: null-hypothesis: null model is the same, covariates are not related to survival

## Exercise:

## Which is the hazard ratio associated with an increase of 10 units in BUN?+

coef of BUN is low because range of BUN is high -\> 1 unit does not change a lot

```{r}
exp(10*0.020706)
```

## -------------------------------------------------------------------------

## Model-based prediction -----------------------------------------------------

## ----------------------------------------------------------------------------

## Estimation of the baseline hazard function

## ------------------------------------------

```{r}
coxph.detail(cox2)
coxph.detail(cox2)$hazard
```

## (Predicted) survival function of an "average" patient

```{r}
cox2$mean
survfit(cox2)
#windows(width = 8)
par(font = 2, font.lab = 4, font.axis = 2, las = 1, mar = c(5, 5, 2, 2))
plot(survfit(cox2), lwd = 3, col = 4, xlab = "Months", ylab = expression(bolditalic(hat(S)(t))))
```

## Patient profiles for prediction

## -------------------------------

```{r}
#create dataframe of 5 columns (each variable)
(newdat <- data.frame(BUN = 25, sex = factor(rep(c("Female", "Male"), 2)), HB = 10, age = 65, BJP = factor(rep(c("No", "Yes"), each = 2))))
```

```{r}
survfit(cox2, newdata = newdat)
summary(survfit(cox2, newdata = newdat))
```

## Graphical respresentation

## -------------------------

```{r}
#windows(width = 8)
par(font = 2, font.lab = 4, font.axis = 2, las = 1)
plot(survfit(cox2, newdata = newdat), col = 1:4, lwd = 3, xlab = "Time [Months]",
     bty = "n", xlim = c(0, 100))
legend("bottomleft", c("Woman without BJP", "Man without BJP",
                       "Woman with BJP", "Man with BJP"),
       col = 1:4, lwd = 2, bty = "n")
title("Model-based prediction of survival function (BUN = 25; HB = 10; Age = 65)")
```

use AFT model to predict a survival probability/time (parametric)

## Residuals and goodness-of-fit of the Cox model -----------------------------

## ----------------------------------------------------------------------------

```{r}
residuals(cox2)
```

## The Schoenfeld residuals:

## Can we assume that the assumption of proportional hazards holds?

## ----------------------------------------------------------------

```{r}
residuals(cox2, "schoenfeld") ## (i) Use of function cox.zph cox.zph(cox2)
```

-   schoenfeld residuals only exists with censored events
-   prportion holds for spec covariate = no correlation between survival time and residuals of a specific covariate

## (ii) Use of function plot.cox.zph

```{r}
#windows(width = 12, height = 7)
par(mfrow = c(2, 3), font = 2, font.lab = 4, font.axis = 2, las = 1, cex.lab = 1.3, cex.axis = 1.2)
plot(cox.zph(cox2), lwd = 2)
```

-   parameter should be constant over time or draw a horizontal line within CI
-   assumption holds
